---
title: "Differential Gene Expression Analysis"
subtitle: "Cortisol response in zebrafish embryos — Hartig et al. 2016"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Setup

Run this chunk first to load all packages and set your working directory.

```{r setup, message=FALSE, warning=FALSE}
library(DESeq2)
library(ggplot2)
library(ggrepel)
library(dplyr)
library(tibble)
library(readr)
library(pheatmap)

# Working directory is set automatically by bulkDGE.Rproj — verify it here:
getwd()  # should end in .../docs/bulkDGE
```

---

## Exercise 1 — Load and Inspect the Data

Load the count matrix and the design table, then answer the questions below.

```{r ex1-load, message=FALSE}
# Load the count matrix
counts_raw <- read_tsv("data/counts.tsv")

# Load the design table
design     <- read_tsv("data/design.txt")

# ── Inspect ──────────────────────────────────────────────────────────────────
dim(counts_raw)    # rows = genes, cols = gene_id + gene_name + 6 samples
head(counts_raw)
colnames(counts_raw)

dim(design)
head(design)
```

**Questions to discuss:**

1. How many genes are in this dataset?
2. What are the six sample names?
3. What column in `design` tells us which treatment group each sample belongs to?

---

## Exercise 2 — Build the DESeqDataSet

DESeq2 needs two things: (1) an integer count matrix with genes as rows and samples as columns, and (2) a `colData` table with sample metadata. Build both and verify they are aligned before proceeding.

```{r ex2-build-dds}
# Step 1: Save gene annotations separately (we'll use these later for gene names)
gene_info <- counts_raw %>%
  select(gene_id, gene_name)

# Step 2: Build the count matrix
#   - gene_id becomes row names
#   - gene_name is removed (not a count column)
#   - Salmon outputs fractional counts; DESeq2 needs integers, so we round
cts <- counts_raw %>%
  select(-gene_name) %>%
  column_to_rownames("gene_id") %>%
  mutate(across(everything(), ~ round(.x) %>% as.integer()))

# Step 3: Build colData from the design table
#   - Use the "name" column (matches count matrix column names) as row names
#   - Set reference level: veh = vehicle (control), cort = cortisol (treatment)
coldata <- design %>%
  select(name, treatment) %>%
  mutate(treatment = factor(treatment, levels = c("veh", "cort"))) %>%
  column_to_rownames("name")

# Step 4: Align column order — colnames(cts) must equal rownames(coldata)
cts <- cts[, rownames(coldata)]
all(colnames(cts) == rownames(coldata))  # must be TRUE before continuing

# Step 5: Build DESeqDataSet
dds <- DESeqDataSetFromMatrix(
  countData = cts,
  colData   = coldata,
  design    = ~ treatment
)
dds
```

**Questions to discuss:**

1. Why do we set `levels = c("veh", "cort")`? What would change if we reversed the order?
2. Why do we check `all(colnames(cts) == rownames(coldata))` before building `dds`?
3. What does a positive `log2FoldChange` mean for this comparison?

---

## Exercise 3 — Filter, Run DESeq2, and Extract Results

Pre-filter very low-count genes, run DESeq2, then extract and explore the differential expression results.

```{r ex3-run-deseq2, message=FALSE}
# Pre-filter: keep genes with at least 10 total counts across all samples
# (reduces computation and improves FDR control)
keep <- rowSums(counts(dds)) >= 10
dds  <- dds[keep, ]
cat("Genes after filtering:", nrow(dds), "\n")

# Run DESeq2
# (estimates size factors, dispersions, and applies the Wald test)
dds <- DESeq(dds)

# Check available contrasts — use the name exactly as shown
resultsNames(dds)

# Extract results for the contrast: cortisol vs vehicle
# Positive log2FoldChange = higher in cortisol
res <- results(dds, name = "treatment_cort_vs_veh")
summary(res)
```

```{r ex3-explore-results}
# Convert to a tidy data frame and add gene names
res_df <- as.data.frame(res) %>%
  rownames_to_column("gene_id") %>%
  left_join(gene_info, by = "gene_id") %>%
  relocate(gene_name, .after = gene_id)

head(res_df)

# How many genes are significant at padj < 0.05?
res_df %>% filter(!is.na(padj), padj < 0.05) %>% nrow()

# Top 10 most significant genes
res_df %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  select(gene_name, log2FoldChange, padj) %>%
  head(10)
```

**Questions to discuss:**

1. Look at the `summary()` output. How many genes are up-regulated? Down-regulated?
2. What is `baseMean`? What does `lfcSE` represent?
3. Why do we use `padj` rather than `pvalue` to call significance?

---

## Exercise 4 — PCA: Do Samples Cluster by Treatment?

Use a variance-stabilizing transform and PCA to check that replicates cluster together and that the treatment groups separate.

```{r ex4-pca, fig.width=6, fig.height=5}
# Variance-stabilizing transform (for visualization only)
vst_data <- vst(dds, blind = FALSE)

# Quick built-in PCA
plotPCA(vst_data, intgroup = "treatment")
```

Now build a polished version with ggplot2:

```{r ex4-pca-ggplot, fig.width=6, fig.height=5}
pca_data <- plotPCA(vst_data, intgroup = "treatment", returnData = TRUE)
pct_var  <- round(attr(pca_data, "percentVar") * 100, 1)

ggplot(pca_data, aes(x = PC1, y = PC2, color = treatment, label = name)) +
  geom_point(size = 4) +
  geom_text_repel(show.legend = FALSE, size = 3) +
  scale_color_manual(values = c(veh = "steelblue", cort = "tomato")) +
  labs(
    title    = "PCA — zebrafish embryo RNA-seq",
    subtitle = "Hartig et al. 2016 (jcoffman001)",
    x        = paste0("PC1: ", pct_var[1], "% variance"),
    y        = paste0("PC2: ", pct_var[2], "% variance"),
    color    = "Treatment"
  ) +
  theme_minimal()
```

**Questions to discuss:**

1. What percentage of variance does PC1 capture? Is that a good sign?
2. Do the three replicates within each group cluster tightly?
3. If one sample appeared as an outlier, what would you do?

---

## Exercise 5 — Volcano Plot: The DE Landscape

Build a volcano plot to visualize all genes at once — significant genes stand out from the background.

```{r ex5-volcano, fig.width=8, fig.height=6}
# Prepare plotting data
plot_df <- res_df %>%
  filter(!is.na(padj), !is.na(log2FoldChange)) %>%
  mutate(
    direction = case_when(
      padj < 0.05 & log2FoldChange > 0 ~ "up in cort",
      padj < 0.05 & log2FoldChange < 0 ~ "down in cort",
      TRUE                             ~ "not significant"
    )
  )

# Label the 15 most significant genes
top_genes <- plot_df %>% arrange(padj) %>% head(15)

ggplot(plot_df, aes(x = log2FoldChange, y = -log10(padj), color = direction)) +
  geom_point(alpha = 0.5, size = 1.2) +
  scale_color_manual(values = c(
    "up in cort"      = "tomato",
    "down in cort"    = "steelblue",
    "not significant" = "grey70"
  )) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_text_repel(
    data = top_genes, aes(label = gene_name),
    size = 3, max.overlaps = 20, show.legend = FALSE
  ) +
  labs(
    title    = "Differential Expression: Cortisol vs Vehicle",
    subtitle = "Zebrafish embryos — Hartig et al. 2016",
    x        = "log2 Fold Change (cort / veh)",
    y        = expression(-log[10] ~ adjusted ~ italic(p)),
    color    = NULL
  ) +
  theme_classic()
```

**Questions to discuss:**

1. What pattern do you see in genes upregulated by cortisol?
2. Are there more up- or down-regulated genes, or roughly symmetric?
3. Look at the gene labels. Do any gene names suggest a biological pathway?

---

## Exercise 6 — MA Plot and Heatmap

An MA plot shows the average expression level (x) against fold change (y). It helps identify whether fold change estimates are biased at low expression levels.

```{r ex6-maplot, fig.width=6, fig.height=5}
plotMA(res, ylim = c(-6, 6), alpha = 0.05,
       main = "MA Plot: Cortisol vs Vehicle")
abline(h = 0, col = "grey40", lty = 2)
```

---

Heatmap of the 50 most significant DE genes, z-scored by row so all genes are on the same color scale.

```{r ex6-heatmap, fig.width=7, fig.height=9}
# Indices of the 50 most significant genes
top50_ids <- res_df %>%
  filter(!is.na(padj)) %>%
  arrange(padj) %>%
  head(50) %>%
  pull(gene_id)

# VST-normalized expression matrix for those genes
mat <- assay(vst_data)[top50_ids, ]

# Replace Ensembl IDs with readable gene names
rownames(mat) <- gene_info$gene_name[match(rownames(mat), gene_info$gene_id)]

# Annotation bar showing treatment group
anno_col <- data.frame(
  Treatment = coldata$treatment,
  row.names = rownames(coldata)
)
anno_colors <- list(Treatment = c(veh = "steelblue", cort = "tomato"))

pheatmap(
  mat,
  scale          = "row",
  annotation_col = anno_col,
  annotation_colors = anno_colors,
  show_rownames  = TRUE,
  fontsize_row   = 7,
  cluster_cols   = TRUE,
  main           = "Top 50 DE Genes — Cortisol vs Vehicle"
)
```

---

## Bonus — Gene Set Enrichment Analysis (GSEA)

GSEA asks whether a predefined biological gene set (such as "inflammatory response") is enriched among genes that change with cortisol treatment, without imposing a significance cutoff.

```{r bonus-gsea-setup, message=FALSE}
library(clusterProfiler)
library(msigdbr)

# MSigDB Hallmark gene sets mapped to Danio rerio via orthology
hallmarks  <- msigdbr(species = "Danio rerio", category = "H")
TERM2GENE  <- hallmarks %>% select(gs_name, ensembl_gene)

# Build a named vector: gene_id → log2FoldChange, sorted descending
rank_vec <- res_df %>%
  filter(!is.na(log2FoldChange), !is.na(gene_id)) %>%
  arrange(desc(log2FoldChange)) %>%
  pull(log2FoldChange, name = gene_id)

head(rank_vec)
tail(rank_vec)
```

```{r bonus-gsea-run, message=FALSE, warning=FALSE}
set.seed(42)
gsea_res <- GSEA(
  geneList     = rank_vec,
  TERM2GENE    = TERM2GENE,
  pvalueCutoff = 0.05,
  verbose      = FALSE
)

as.data.frame(gsea_res) %>%
  select(ID, NES, pvalue, p.adjust) %>%
  arrange(p.adjust)
```

```{r bonus-gsea-plot, fig.width=9, fig.height=6}
dotplot(gsea_res, x = "NES", showCategory = 15, label_format = 50) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(
    title    = "GSEA: Cortisol vs Vehicle",
    subtitle = "MSigDB Hallmark gene sets · Danio rerio"
  )
```
