---
title: "Spatial Transcriptomics Analysis"
subtitle: "10x Genomics Visium Mouse Brain: Seurat Spatial Vignette"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Setup

Run this chunk first to load all packages and set your working directory.

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(SeuratData)
library(patchwork)
library(dplyr)
library(ggplot2)

# Working directory is set automatically by spatial.Rproj, verify it here:
getwd()  # should end in .../docs/spatial
```

---

## Part 1: Loading Spatial Data & Pre-processing

## Exercise 1: Load the Visium Dataset and Inspect the Seurat Object

```{r ex1-load}
InstallData("stxBrain")
brain <- LoadData("stxBrain", type = "anterior1")
brain

names(brain@images)
```

---

## Exercise 2: Visualize QC Metrics on the Violin Plot and on the Tissue Itself

```{r ex2-qc, fig.height=6, fig.width=12}
plot1 <- VlnPlot(brain, features = "nCount_Spatial", pt.size = 0.1) + NoLegend()
plot2 <- SpatialFeaturePlot(brain, features = "nCount_Spatial") + theme(legend.position = "right")
plot1 + plot2
```

---

## Exercise 3: Run SCTransform and Visualize Marker Genes on the Tissue

```{r ex3-sctransform, fig.height=6, fig.width=12}
brain <- SCTransform(brain, assay = "Spatial", verbose = FALSE)

SpatialFeaturePlot(brain, features = c("Hpca", "Ttr"))
```

Try adjusting `alpha` to see how it changes the plot:

```{r ex3-alpha, fig.height=6, fig.width=12}
SpatialFeaturePlot(brain, features = c("Hpca", "Ttr"), alpha = c(0.1, 1))
```

---

## Part 2: Post-processing

## Exercise 4: Run PCA and Choose Your Number of PCs

Inspect the elbow plot, then set `n_pcs` to whatever cutoff you decide on. Every remaining step uses `dims = 1:n_pcs`.

```{r ex4-pca-elbow, fig.height=6, fig.width=10, results='hide'}
brain <- RunPCA(brain, assay = "SCT", verbose = FALSE)
ElbowPlot(brain, ndims = 30)

n_pcs <- 30   # pick your own cutoff based on the elbow plot above
```

---

## Exercise 5: Cluster Spots and Compare UMAP to the Tissue

```{r ex5-cluster-umap, fig.height=7, fig.width=13}
brain <- FindNeighbors(brain, reduction = "pca", dims = 1:n_pcs)
brain <- FindClusters(brain, verbose = FALSE)
brain <- RunUMAP(brain, reduction = "pca", dims = 1:n_pcs)

p1 <- DimPlot(brain, reduction = "umap", label = TRUE)
p2 <- SpatialDimPlot(brain, label = TRUE, label.size = 3)
p1 + p2
```

Isolate a single cluster's footprint on the tissue:

```{r ex5-highlight-cluster, fig.height=6, fig.width=8}
my_explore_cluster <- 0   # try a few different cluster numbers here

SpatialDimPlot(brain, cells.highlight = WhichCells(brain, idents = my_explore_cluster))
```

---

## Part 3: Spatially Variable Genes & Region Annotation Assignment

## Exercise 6: Find Markers Between Clusters, and Find Spatially Variable Genes

```{r ex6-de-markers, fig.height=6, fig.width=12}
# Pick two clusters that looked spatially interesting to you in Exercise 5
de_markers <- FindMarkers(brain, ident.1 = 1, ident.2 = 2)
SpatialFeaturePlot(object = brain, features = rownames(de_markers)[1:3], alpha = c(0.1, 1))
```

```{r ex6-spatially-variable, fig.height=8, fig.width=13}
brain <- FindSpatiallyVariableFeatures(
  brain,
  assay = "SCT",
  features = VariableFeatures(brain)[1:1000],
  selection.method = "moransi"
)

top.features <- head(SpatiallyVariableFeatures(brain, method = "moransi"), 6)
SpatialFeaturePlot(brain, features = top.features, ncol = 3, alpha = c(0.1, 1))
```

---

## Exercise 7: Research Your Assigned Cluster's Region Identity

Your instructor will assign you one cluster number from **your own** results.

```{r ex7-my-cluster-markers}
my_cluster <- 0   # replace with the cluster number your instructor assigned you

FindMarkers(brain, ident.1 = my_cluster) %>%
  filter(pct.1 > 0.25) %>%
  arrange(desc(avg_log2FC)) %>%
  head(15)
```

Once you've picked 3-5 candidate marker genes from that table and researched them, sanity-check them against your cluster's location on the tissue:

```{r ex7-my-cluster-plots, fig.height=6, fig.width=12}
my_genes <- c("GENE1", "GENE2", "GENE3")  # replace with your chosen marker genes

SpatialFeaturePlot(brain, features = my_genes, alpha = c(0.1, 1))
SpatialDimPlot(brain, cells.highlight = WhichCells(brain, idents = my_cluster))
```

---

## Finalize the Annotation as a Class

After everyone presents their assigned cluster, fill in the agreed-upon labels in cluster order and apply them together.

```{r finalize-annotation, fig.height=6, fig.width=9}
new.cluster.ids <- c("cluster0_label", "cluster1_label", "cluster2_label")  # extend/edit to match your class's clusters
names(new.cluster.ids) <- levels(brain)
brain <- RenameIdents(brain, new.cluster.ids)

SpatialDimPlot(brain, label = TRUE, label.size = 3)
```
