---
title: "Single Cell RNA-seq Analysis"
subtitle: "PBMC 3k: 10x Genomics / Seurat Guided Clustering Tutorial"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Setup

Run this chunk first to load all packages and set your working directory.

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(patchwork)
library(dplyr)
library(ggplot2)

# Working directory is set automatically by scRNAseq.Rproj, verify it here:
getwd()  # should end in .../docs/scRNAseq
```

---

## Part 1: QC & Pre-processing

## Exercise 1: Load the Data and Build a Seurat Object

```{r ex1-load}
pbmc.data <- Read10X(data.dir = "data/pbmc3k/filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc
```

---

## Exercise 2: Score Mitochondrial Content and Visualize QC Metrics

```{r ex2-qc, fig.height=7, fig.width=13}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
summary(pbmc$percent.mt)

# Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# FeatureScatter visualizes feature-feature relationships
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

---

## Exercise 3: Choose Your Own QC Thresholds and Filter

Look at the plots from Exercise 2 and decide your own cutoffs. There is no single correct answer here; record your reasoning in the workshop questions.

```{r ex3-filter}
min_features   <- 200    # lower bound on nFeature_RNA, pick your own
max_features   <- 2500   # upper bound on nFeature_RNA (your doublet proxy), pick your own
max_percent_mt <- 5      # upper bound on percent.mt, pick your own

pbmc <- subset(
  pbmc,
  subset = nFeature_RNA > min_features & nFeature_RNA < max_features & percent.mt < max_percent_mt
)
pbmc
```

---

## Exercise 4: Re-Visualize QC Metrics After Filtering

Run the *same* plotting calls from Exercise 2 again, now on your filtered object, and compare.

```{r ex4-qc-recheck, fig.height=7, fig.width=13}
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

---

## Exercise 5: Normalize and Find Variable Features

```{r ex5-normalize-variable, fig.height=5, fig.width=10}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 1e4)
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)
top10

plot1 <- VariableFeaturePlot(pbmc)
LabelPoints(plot = plot1, points = top10, repel = TRUE)
```

---

## Exercise 6: Scale the Data

```{r ex6-scale, results='hide'}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

---

## Part 2: Post-processing

## Exercise 7: Run and Visualize PCA

```{r ex7-pca, fig.height=8, fig.width=10, results='hide'}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))

print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
DimPlot(pbmc, reduction = "pca") + NoLegend()
```

---

## Exercise 8: Choose Your Number of PCs

Inspect the elbow plot, then set `n_pcs` to whatever cutoff you decide on. Every remaining step uses `dims = 1:n_pcs`, so this single choice propagates through the rest of your analysis.

```{r ex8-elbow, fig.height=6, fig.width=10}
ElbowPlot(pbmc, ndims = 30)

n_pcs <- 10   # pick your own cutoff based on the elbow plot above
```

---

## Exercise 9: Build the KNN Graph and Cluster with Louvain

```{r ex9-cluster}
pbmc <- FindNeighbors(pbmc, dims = 1:n_pcs)
pbmc <- FindClusters(pbmc, resolution = 0.5)

# Look at cluster IDs of the first 5 cells
head(Idents(pbmc), 5)
table(Idents(pbmc))
```

---

## Exercise 10: Run UMAP

```{r ex10-umap, fig.height=8, fig.width=8}
pbmc <- RunUMAP(pbmc, dims = 1:n_pcs)
DimPlot(pbmc, reduction = "umap")
```

---

## Part 3: Cluster Marker Research Assignment

## Exercise 11: Find All Cluster Marker Genes

```{r ex11-markers}
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)

pbmc.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 10) %>%
  ungroup()
```

---

## Exercise 12: Annotate Your Assigned Cluster

Your instructor will assign you one cluster number from **your own** results.

```{r ex12-my-cluster-markers}
my_cluster <- 0   # replace with the cluster number your instructor assigned you

FindMarkers(pbmc, ident.1 = my_cluster) %>%
  filter(pct.1 > 0.25) %>%   # expressed in a meaningful fraction of the cluster, not just a couple of cells
  arrange(desc(avg_log2FC)) %>%
  head(15)
```

Once you've picked 3-5 candidate marker genes from that table and researched them, sanity-check them against your cluster:

```{r ex12-my-cluster-plots, fig.height=5, fig.width=10}
my_genes <- c("GENE1", "GENE2", "GENE3")  # replace with your chosen marker genes

VlnPlot(pbmc, features = my_genes)
FeaturePlot(pbmc, features = my_genes)
```

---

## Finalize the Annotation as a Class

After everyone presents their assigned cluster, fill in the agreed-upon labels in cluster order and apply them together.

```{r finalize-annotation, fig.height=5, fig.width=9}
new.cluster.ids <- c("cluster0_label", "cluster1_label", "cluster2_label")  # extend/edit to match your class's clusters
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)

DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```
