---
title: "Single Cell RNA-seq Analysis"
subtitle: "PBMC 3k — 10x Genomics / Seurat Guided Clustering Tutorial"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
---

## Setup

Run this chunk first to load all packages and set your working directory.

```{r setup, message=FALSE, warning=FALSE}
library(Seurat)
library(patchwork)
library(dplyr)
library(ggplot2)

# Working directory is set automatically by scRNAseq.Rproj — verify it here:
getwd()  # should end in .../docs/scRNAseq
```

---

## Part 1 — QC & Pre-processing

## Exercise 1 — Load the Data and Build a Seurat Object

```{r ex1-load}
pbmc.data <- Read10X(data.dir = "data/pbmc3k/filtered_gene_bc_matrices/hg19/")
pbmc <- CreateSeuratObject(counts = pbmc.data, project = "pbmc3k", min.cells = 3, min.features = 200)
pbmc
```

---

## Exercise 2 — Score Mitochondrial Content and Visualize QC Metrics

```{r ex2-qc, fig.height=7, fig.width=13}
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")
summary(pbmc$percent.mt)

# Visualize QC metrics as a violin plot
VlnPlot(pbmc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

# FeatureScatter visualizes feature-feature relationships
plot1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```

---

## Exercise 3 — Filter Low-Quality Cells and Doublets

```{r ex3-filter}
pbmc <- subset(pbmc, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 5)
pbmc
```

---

## Exercise 4 — Normalize and Find Variable Features

```{r ex4-normalize-variable, fig.height=5, fig.width=10}
pbmc <- NormalizeData(pbmc, normalization.method = "LogNormalize", scale.factor = 1e4)
pbmc <- FindVariableFeatures(pbmc, selection.method = "vst", nfeatures = 2000)

# Identify the 10 most highly variable genes
top10 <- head(VariableFeatures(pbmc), 10)
top10

plot1 <- VariableFeaturePlot(pbmc)
LabelPoints(plot = plot1, points = top10, repel = TRUE)
```

---

## Exercise 5 — Scale the Data

```{r ex5-scale, results='hide'}
all.genes <- rownames(pbmc)
pbmc <- ScaleData(pbmc, features = all.genes)
```

---

## Part 2 — Post-processing

## Exercise 6 — Run and Visualize PCA

```{r ex6-pca, fig.height=8, fig.width=10, results='hide'}
pbmc <- RunPCA(pbmc, features = VariableFeatures(object = pbmc))

print(pbmc[["pca"]], dims = 1:5, nfeatures = 5)
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
DimPlot(pbmc, reduction = "pca") + NoLegend()

# Elbow plot — how many PCs carry real signal?
ElbowPlot(pbmc)
```

---

## Exercise 7 — Build the KNN Graph and Cluster with Louvain

```{r ex7-cluster}
pbmc <- FindNeighbors(pbmc, dims = 1:10)
pbmc <- FindClusters(pbmc, resolution = 0.5)

# Look at cluster IDs of the first 5 cells
head(Idents(pbmc), 5)
table(Idents(pbmc))
```

---

## Exercise 8 — Run UMAP

```{r ex8-umap, fig.height=8, fig.width=8}
pbmc <- RunUMAP(pbmc, dims = 1:10)
DimPlot(pbmc, reduction = "umap")
```

---

## Bonus — Marker Genes and Cell Type Annotation

```{r bonus-markers}
pbmc.markers <- FindAllMarkers(pbmc, only.pos = TRUE)

pbmc.markers %>%
  group_by(cluster) %>%
  dplyr::filter(avg_log2FC > 1) %>%
  slice_head(n = 5) %>%
  ungroup()
```

```{r bonus-annotate, fig.height=5, fig.width=9}
new.cluster.ids <- c("Naive CD4 T", "CD14+ Mono", "Memory CD4 T", "B", "CD8 T",
                      "FCGR3A+ Mono", "NK", "DC", "Platelet")
names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)

DimPlot(pbmc, reduction = "umap", label = TRUE, pt.size = 0.5) + NoLegend()
```
